Revealing Transient Strain in Geodetic Data with the Radial Basis Function Finite Difference Method

### 2017-06-01 HinesHetlandGPR.tex
added more crap

\subsection{Spatial and temporal covariance functions}
We must choose an appropriate Gaussian process describing transient displacements $u$ and GNSS noise $\epsilon$. We assume that $u$ and $\epsilon$ both have zero mean, which leaves for use to select a parameterization for the covariance functions $X$, $T$, and $C_d$.  In this section, we list the covariance functions which are considered in this paper and those which could be of use in future studies. In the following discussion $\alpha$ and $\beta$ are hyperparameters. There are numerous strategies for selecting these hyperparameters (see \citet{Cressie1992}) and we save a discussion on choosing the hyperparameters for Section X.

The squared exponential covariance function,
\begin{equation}\label{eq:SE}
C(x,x') = \alpha^2 \exp\left(\frac{-||x - x'||_2^2}{2 \beta^2}\right),
\end{equation}
is commonly used in Kriging \citep[e.g,][]{Cressie1992} and Gaussian process regression \citep[e.g.,][]{Rasmussen2006}.  The squared exponential is a valid (i.e. positive definite) covariance function for any number of spatial dimensions, and it describes an isotropic (i.e. $C(x,x')$ can be expressed as $\hat{C}(||x - x'||_2)$) Gaussian processes with realizations that are infinitely differentiable.  \citet{Kato1998} and \cite{El-Fiky1999} demonstrated that eq. (\ref{eq:SE}) is an appropriate covariance model for describing long-term tectonic strain rates for Japan.

The Wendland class of covariance functions \citep{Wendland2005} are positive definite in $\mathbb{R}^d$, and they describes an isotropic Gaussian process with realizations that can be differentiated $k$ times. The form of the covariance function depends on the choice of $d$ and $k$. Wendland covariance functions have compact support, which is a particularly useful property because we can exploit the sparsity of the the corresponding covariance matrices. In this study we use Wendland covariance functions to describe the temporal component of transient displacements. Therefore, we only require that $d=1$ and $k=1$. The corresponding Wendland covariance function is 
\begin{equation}\label{eq:Wendland}
C(x,x') = \alpha^2\left(1 - \frac{|x - x'|}{\beta}\right)^3_+ \left(\frac{3|x - x'|}{\beta} + 1\right), \ \ \ x,x' \in \mathbb{R}^1,
\end{equation}
where
\begin{equation}
(x)_+ = 
\begin{cases}
x, \ \ \ x > 0 \\
0, \ \ \ \mathrm{otherwise}.
\end{cases}
\end{equation}

We next consider the covariance function which describes a first-order Gauss-Markov (FOGM) process. A FOGM process is a solution to the stochastic differential equation
\begin{equation}
\dot{u}(x) + \alpha u(x) = \beta w(x), \ \ \ x,x' \in \mathbb{R}^1,
\end{equation}
where $w(x)$ is white noise with unit variance.  The solution is a Gaussian process with zero mean and covariance
% This needs to be scaled to match my work
\begin{equation}\label{eq:FOGM}
C(x,x') = \frac{\beta^2}{2\alpha}\exp\left(-\alpha|x - x'|\right), \ \ \ x,x' \in \mathbb{R}^1. 
\end{equation}
FOGM processes are isotropic and realizations are not differentiable.  The FOGM model has been used to describe temporally correlated noise in EDM data \citep{Langbein2004} and GNSS data \citep[e.g.,][]{Langbein2008}.

One last covariance function which is worth mentioning, even though we do not use it in this paper, is the periodic covariance function \citep{Mackay1998}, 
\begin{equation}\label{eq:Periodic}
C(x,x') = \alpha^2 \exp\left(\frac{-\sin(\frac{\pi}{\lambda} |x - x'|)^2}{2 \beta^2} \right), \ \ \ x,x' \in \mathbb{R}^1.
\end{equation}
The periodic covariance function describes an isotropic Gaussian process with realizations that have period $\lambda$.  The length-scale of features within each period is controlled by $\beta$. In terms of GNSS data, a periodic Gaussian process can be used to describe seasonal deformation. The more common method of treating seasonal deformation is to model it as a linear combination of annual and semi-annual sinusoids \citep[e.g.,][]{Dong2002}. We believe there is potential to use periodic Gaussian processes to explore the statistical properties of seasonal signals; however, that is outside the scope of this paper.

Brownian motion and integrated Brownian motion are two Gaussian processes which are frequently used in the geodetic literature.  Brownian motion, which has zero mean and covariance function
\begin{equation}\label{eq:BM}
C(x,x') =\alpha^2 \min(x,x'), \ \ \ x,x' \in \mathbb{R}^1 \ \mathrm{and} \ \ x,x' \geq 0, 
\end{equation} 
is commonly used to describe noise resulting from the instability of geodetic monuments \citep[e.g.,][]{Wyatt1982,Wyatt1989,Langbein1997}. Integrated Brownian motion, which has zero mean and covariance function
\begin{align}\label{eq:IBM}
C(x,x') &= \int_0^x \int_0^{x'} \alpha^2 \min(u,u') \,du'\,du \\
        &= \frac{\alpha^2}{2}\min(x,x')^2 \left(\max(x,x') - \frac{1}{3}\min(x,x')\right), \ \ \ x,x' \in \mathbb{R}^1 \ \mathrm{and} \ \ x,x' \geq 0,
\end{align}
has been used as a nonparametric model for the time dependence of geophysical signals \citep[e.g.,][]{Segall1997,McGuire2003,Ohtani2010,Hines2016}. It should be noted that Brownian motion and integrated Brownian motion have a reference time at which the Gaussian process is exactly zero. For some geophysical signals, such as postseismic deformation, it is reasonable to have such a reference time. On the other hand, if there is no sense of a start time for the geophysical process, then Brownian motion or integrated Brownian motion may not be appropriate, and one of the isoptropic Gaussian processes described above should be used.

\section{Application Cascadia Slow Slip Events}\label{sec:Cascadia}

We demonstrate the method for estimating transient strain described in the previous section by applying it to GNSS data in Cascadia. We are mainly interested in identifying transient strain resulting from slow slip events (SSEs) \citep[e.g.,][]{Dragert2001}. We show that Gaussian processes are an effective tool for (1) detecting SSEs and (2) understanding how the strain from SSEs are contributing to the strain accumulated over the seismic cycle. Our primary motivation for detecting SSEs is because there have been several documented cases of SSEs triggering megathrust earthquakes \citep{Roeloffs2006}.  In Cascadia, there is less of a need to detect SSEs with GNSS data because the SSEs are associated with a spike in seismic tremor, which is actively monitored by the Pacific Northwest Seismic Network (PNSN) \citep{Wech2010}. However, SSEs have been observed in other subduction zones without any associated tremor \citep{Schwartz2007}, and so there is still utility in developing method for detecting SSEs with GNSS data. Furthermore, by comparing our estimated transient strain in Cascadia with the tremor catalog, we can verify the accuracy of method for detecting SSEs.

\subsection{Noise model}
In this section we discuss our choice for the noise covariance function $C_\epsilon$. We will assume that the noise consists of uncorrelated noise with standard deviation given by the formal data uncertainties and potentially spatially and/or temporally correlated noise. There have been numerous studies on temporally correlated noise in GNSS data \citep[e.g.,]{Zhang1997,Mao1999,Williams2004,Langbein2008}. These studies have found some combination of Brownian motion or flicker noise best describes the temporally correlated noise in GNSS data. Brownian motion has also been used to describe error resulting from an unstable geodetic monument \citep[e.g.,][]{Wyatt1982,Wyatt1989}. We assume that temporally correlated noise can be described by a FOGM stochastic model, noting that the Brownian motion is a degenerate case of a FOGM process.

To determine the appropriate hyperparameters for the FOGM model, $\alpha$ and $\beta$, we first select 38 stations in in Washington and Oregon which are east of $121^\circ$ W, where we can reasonably expect no transient signal from SSEs (i.e. $u = 0$). We then find $\alpha$ and $\beta$ for each station timeseries with the Restricted Maximum Likelihood (REML) method \cite[e.g.,][]{Harville1974,Cressie1992} and Hines 2017. The REML method finds the hyperparameters which maximize the probability of drawing $\bm{d}_*$ from $\bm{d}$. We use the REML method over the maximum likelihood (ML) method \citep[e.g.,][]{Langbein1997} because the REML method properly accounts for the improper prior that we assigned to $\bm{a}$. The distribution of inferred $\alpha$ and $\beta$ are shown in Figure X. The amplitude of FOGM noise for the east and north components is notable low and are unimodally clustered around 0.5 mm/yr$^{0.5}$. The corresponding estimates of $\alpha$ tend to cluster around 0 yr$^{-1}$, which is the degenerate case where FOGM behaves likes Brownian motion.  The amplitude of FOGM noise for the vertical component is significantly larger with a median value of 13.8 mm/yr$^{0.5}$. The inferred values for alpha are higher for the vertical component with a median value of 1.87 yr$^{-1}$. In Figure X, we use the median values of $\alpha$ and $\beta$ to generate two random samples of FOGM noise. The samples span seven years and over these seven years the east and north samples drift by about 1 mm. In the context of detecting SSEs, which produce several mm's of surface displacement on the timescale of weeks to months, the estimated FOGM noise for the east and north component is negligible. In contrast, the estimated FOGM noise for the vertical component is significantly larger than the signal we would expect from SSEs. We suspect that the higher amplitude for the FOGM model in the vertical component is accommodating for deficiencies in our rather simple seasonal model. Based on this analysis, we henceforth ignore temporally correlated noise in the east and north component because of its low amplitude, and we do not estimate $u$ for vertical deformation because of the low signal to noise ratio.

Another significant source of noise in GNSS data is common mode error \citep[e.g.,][]{Wdowinski1997}, which is noise that is highly spatially correlated. When not accounted for, common mode error manifests itself as spatially uniform undulations in $\hat{u}$. However, we are primarily interested in estimating strain which is insensitive to common mode error. We therefore do not include common mode error in our noise model.

Our noise is just white.

\subsection{Transient displacement model}

@@

Explain strain concentration with this:
Slow slip event depths from \citet{Dragert2001}, \citet{Wech2009}, \citet{Schmidt2010}, \citet{Bartlow2011} show slip concentrated at depths from 30 to 50 km detph.

>>>>

GIVE A DEMONSTRATION HERE

\section{Application Cascadia Slow Slip Events}\label{sec:Cascadia}
We demonstrate the method for estimating transient strain described in the previous section by applying it to GNSS data in Cascadia. We are mainly interested in identifying transient strain resulting from slow slip events (SSEs) \citep[e.g.,][]{Dragert2001}. We show that Gaussian processes are an effective tool for (1) detecting SSEs and (2) understanding how the strain from SSEs are contributing to the strain accumulated over the seismic cycle. Our primary motivation for detecting SSEs is because there have been several documented cases of SSEs triggering megathrust earthquakes \citep{Roeloffs2006}.  In Cascadia, there is less of a need to detect SSEs with GNSS data because the SSEs are associated with a spike in seismic tremor, which is actively monitored by the Pacific Northwest Seismic Network (PNSN) \citep{Wech2010}. However, SSEs have been observed in other subduction zones without any associated tremor \citep{Schwartz2007}, and so there is still utility in developing method for detecting SSEs with GNSS data. Furthermore, by comparing our estimated transient strain in Cascadia with the tremor catalog, we can verify the accuracy of method for detecting SSEs.

\subsection{Noise model}
In this section we discuss our choice for the noise covariance function $C_\epsilon$. We will assume that the noise consists of an uncorrelated component with standard deviation given by the formal data uncertainties, $\bm{\sigma}$, and a spatially and/or temporally correlated component. There have been numerous studies on temporally correlated noise in GNSS data \citep[e.g.,]{Zhang1997,Mao1999,Williams2004,Langbein2008}. In these studies, temporally correlated noise was described with some combination of Brownian motion, first-order Gauss-Markov (FOGM) processes, and/or flicker noise. There is some physical justification for using Brownian motion as a noise model because it accurately describes the power spectrum of motion resulting from instability in geodetic monuments \citep[e.g.,][]{Wyatt1982,Wyatt1989,Langbein1997}. We describe temporally correlated noise as a FOGM process, noting that Brownian motion is a degenerate case of a FOGM process. A FOGM process is a solution to the stochastic differential equation
\begin{equation}
\dot{v}(t) + \alpha v(t) = \beta w(t),
\end{equation}
where $w(t)$ is white noise with unit variance. The solution is an isotropic Gaussian process with zero mean and covariance function
\begin{equation}\label{eq:FOGM}
C_v(t,t') = \frac{\beta^2}{2\alpha}\exp\left(-\alpha|t - t'|\right). 
\end{equation}
The FOGM model degenerates to Brownian motion under the condition that $\alpha = 0$ and $v(0) = 0$.

To determine the appropriate hyperparameters for the FOGM model, $\alpha$ and $\beta$, we first select 38 stations in Washington and Oregon which are east of $121^\circ$ W, where we can reasonably expect no transient signal from SSEs (i.e. $u = 0$). We then find $\alpha$ and $\beta$ for each station timeseries with the Restricted Maximum Likelihood (REML) method \cite[e.g.,][]{Harville1974,Cressie1992} and Hines 2017. The REML method finds the hyperparameters which maximize the probability of drawing $\bm{d}_*$ from $\bm{d}$. We use the REML method over the maximum likelihood (ML) method \citep[e.g.,][]{Langbein1997} because the REML method accounts for the improper prior that we assigned to $\bm{a}$ without biasing the hyperparameter estimates. The distribution of inferred $\alpha$ and $\beta$ are shown in Figure X. The amplitude of FOGM noise for the east and north components is notable low and are clustered around 0.5 mm/yr$^{0.5}$. The corresponding estimates of $\alpha$ tend to cluster around 0 yr$^{-1}$, suggesting that Brownian motion would have also been an appropriate noise model.  The amplitude of FOGM noise for the vertical component is significantly larger with a median value of 13.8 mm/yr$^{0.5}$. The inferred values for alpha are higher for the vertical component with a median value of 1.87 yr$^{-1}$. In Figure X, we use the median values of $\alpha$ and $\beta$ to generate two random samples of FOGM noise. The samples span seven years and over these seven years the east and north samples drift by about 1 mm. In the context of detecting SSEs, which produce several mm's of surface displacement on the timescale of weeks, the estimated FOGM noise for the east and north component is negligible. In contrast, the estimated FOGM noise for the vertical component is significantly larger than the signal we would expect from SSEs. We suspect that the higher amplitude for the FOGM model in the vertical component is accommodating for deficiencies in our rather simple seasonal model. Based on this analysis, we henceforth ignore temporally correlated noise in the east and north component because of its low amplitude, and we do not estimate $u$ for vertical deformation because of the low signal to noise ratio.

Another significant source of noise in GNSS data is common mode error \citep[e.g.,][]{Wdowinski1997,Dong2006}, which is noise that is highly spatially correlated. When not accounted for, common mode error manifests itself as spatially uniform undulations in $\hat{u}$. However, we are primarily interested in estimating strain which is insensitive to common mode error. We therefore do not include common mode error in our noise model.

For our estimation of of transient strain resulting from SSEs, we then simply assume that $\epsilon$ consists of only uncorrelated noise with covariance $C_\epsilon(\bm{y},\bm{y}') = \mathrm{diag}(\sigma^2)$.

\subsection{Transient displacement model}
We next discuss our prior model for transient displacements, and, specifically, our choice for the covariance functions $X(x,x')$ and $T(t,t')$. For $X$, we use the squared exponential (SE) covariance function,
\begin{equation}\label{eq:SE}
X(x,x') = \alpha^2 \exp\left(\frac{-||x - x'||_2^2}{2 \beta^2}\right).
\end{equation}
The SE covariance function is commonly used in Kriging \citep[e.g,][]{Cressie1992} and Gaussian process regression \citep[e.g.,][]{Rasmussen2006}.  The SE is a valid (i.e. positive definite) covariance function for any number of spatial dimensions, and it describes an isotropic Gaussian processes with realizations that are infinitely differentiable. In terms of geodetic applications, \citet{Kato1998} and \cite{El-Fiky1999} demonstrated that the SE is an appropriate covariance model for describing long-term tectonic strain rates in Japan.

We consider three potential covariance functions to describe the temporal covariance of $u$.  First, we consider the one-dimensional SE covariance function, simply because the SE describes a good general purpose Gaussian process. Second, we consider integrated Brownian motion (IBM). IBM has zero mean and its covariance function can be found by integrating the covariance function for Brownian motion as
\begin{align}\label{eq:IBM}
T(t,t') &= \int_0^t \int_0^{t'} \alpha^2 \min(\tau,\tau') \,d\tau'\,d\tau \\
        &= \frac{\alpha^2}{2}\min(t,t')^2 \left(\max(t,t') - \frac{1}{3}\min(t,t')\right), \ \ \ t,t' \geq 0.
\end{align}
IBM has been used, in the context of Kalman filtering, as a nonparametric model for the time dependence of geophysical signals \citep[e.g.,][]{Segall1997,McGuire2003,Ohtani2010,Hines2016}. It should be noted that IBM has a reference time at which the Gaussian process is exactly zero. For some geophysical signals, such as postseismic deformation, it is appropriate to have this reference time. However, if we are interesting in detecting transient events, where there is no known start time, then the IBM model may not be appropriate and an isotropic Gaussian process should be preferred. Alternatively we can assume that the start date for the IBM process is long before the first data obvservation; however this can introduce numerical difficulties because the covariance matrix will become ill-conditioned. Third, we consider the Wendland class of covariance functions \citep{Wendland2005}. Wendland functions are positive definite in $\mathbb{R}^d$, and they describes an isotropic Gaussian process with realizations that can be differentiated $k$ times. The form of the covariance function depends on the choice of $d$ and $k$. Wendland covariance functions have compact support, which is a particularly useful property because we can exploit the sparsity of the the corresponding covariance matrices. To describe the temporal covariance of $u$, we only require that $d=1$ and $k=1$. The corresponding Wendland covariance function is 
\begin{equation}\label{eq:Wendland}
T(t,t') = \alpha^2\left(1 - \frac{|t - t'|}{\beta}\right)^3_+ \left(\frac{3|t - t'|}{\beta} + 1\right), 
\end{equation}
where
\begin{equation}
(t)_+ = 
\begin{cases}
t, \ \ \ t > 0 \\
0, \ \ \ \mathrm{otherwise}.
\end{cases}
\end{equation}

We next find the optimal hyperparameters for $X$ and each of the three candidate functions for $T$. We want to find the hyperparameters which best describes the spatial and temporal statistics of ground deformation resulting from SSEs. We focus on describing the SSEs in the Puget Sound region, which regularly occur with a frequency of about 14 months \citep{Brudzinski2007}.

To describe these events we use the set of GNSS station shown in Figure X. We use the GNSS data that has been made publicly available through UNAVCO \citep{Herring2016}. We first remove timeseries jumps based on the compiled GNSS offsets from blewitt. We then use the data editing algorithm to remove outliers. We find the data editing algorithm is particularly useful at removing outliers for stations at higher altitudes, which can be adversely affected by snow \citep{Lisowski2008}. For our data editing algorithm we use X for the prior and we give an example in figure X.

We use the REML method to find the hyperparameters that best describe SSE events which occurred since 2010, corresponding to the time that coverage improved due to the PBO network.  We identify the times of seven SSEs using tremor records from \cite{Wech2010}, and we break the spatio-temporal GNSS data into four month long subsets roughly centered on the time of each SSE. For each data subset use the REML method to find the hyperparameters for $T$ and $X$ that best describe deformation from the SSE. Our motivation analyzing subsets of the data is two-fold: (1) it keeps the analysis computationally tractible and (2) the distribution of estimated hyperparameters will indicates whether we can use a single covariance model to describe all SSEs. Our hyperparameter estimates are summarized in Table X.

We next want to identify which covariance function for $T$ best describes the SSEs. We could use the REML likelihoods to compare different models, similar to \citep{Langbein2004}. However, the maximum likelihoods do not vary by more than a factor of X between the models, which does not definitively identify the best model. We then compare the posterior estimates $\hat{u}$ for each model to the observations to see if any model is noticeably better at describing the data. We show $\hat{u}$ using the three different function for $T$ in Figure X. It is clear that neither model offer any distinct improvement over the others. For most stations $\hat{u}$ appears to accurately capture the deformation from SSEs; however, for some stations which are closer to Puget Sound such as X and X, the rate of deformation is underestimated by the best fitting model. We could have manually specified a prior. We choose the Wendland function to describe $T$ because evaluating $T$ will yield sparse covariance matrices. We utilize the sparsity of the covariance matrices with the sparse matrix library CHOLMOD \citep{Chen2008}, which significantly expands the amount of data which we are able to tractably analyze with the method described in Section X.

\subsection{Transient Strain Rates}

Having established noise model and a prior for transient displacements, we calculate the transient strain rates in the Puget Sound region from eq. (\ref{eq:StrainRate}) over the time period 2010-01-01 to 2017-05-15. Maybe this is the best place to mention the prior for editing the data. Our solution is shown in Figures X through X. Figures X through X show the estimates $\dot{\bm{\epsilon}}$ complete with uncertainties at select times over the course of the Winter 2016 SSE. Figure X, which shows strain rate at the height of the SSE, makes the implications of SSEs for seismic hazard clear.  The SSE causes compression in the Olympic Peninsula and extension in East of Puget Sound. The SSE are thus concentrating strain energy trench-ward, and pushing the subduction zone closer to failure. This is not novel insight because the same conclusions has been drawn from fault slip models which show that SSEs tend to be just at the transition between locked and decoupled. Slow slip event depths from \citet{Dragert2001}, \citet{Wech2009}, \citet{Schmidt2010}, \citet{Bartlow2011} show slip concentrated at depths from 30 to 50 km detph. What is novel is that our estimates of strain rate from SSEs are nonparametric and less likely to be biased by systematic errors such as errors in the fault model. Are uncertainties are also likely to be more conservative since we do not make any dubious assumptions. Thus we can be more confident in our assessment of how much strain is accumulating from SSEs.

To demonstrate the effectiveness of our method for detecting SSEs, we show a timeseries of estimated strain components at the position indicated in figure X.  We also compare the strain rate magnitude and the frequency of tremors in this region.  The strain rate magnitudes have been normalized by its estimated uncertainty and so strain magnitudes less than ${\sim}$2 can be considered to be part of the background noise. The only time that the normalized strain rate magnitude exceeds 2 is when there is a corresponding peak in tremor.  We are able to accurately detect the seven SSEs in over this period in addition to the inter-SSE events such as the one at X.

made in ,  is accumulating to the trench perpendicular compression releases At the height of the SSE, the implications of

@@

Explain strain concentration with this:
